---
title: "Exploratory Data Visuals"
author: "Rosie Wu"
date: "2025-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import libraries and Combined Dataset
```{r}
library(here)
library(tidyverse)
library(gtsummary)
library(broom)
library(ggplot2)
library(dplyr)

main_df <- read.csv("Rosie's Docs/Combined Final Dataframe.csv")

```

# Exploratory Data Visuals
## Dependent Variables
```{r PM2.5 Value Distribution}
ggplot(main_df, aes(x = PM.2.5)) +
  geom_histogram(bins = 10, fill = "blue", color = "black") +
  ggtitle("Distribution of PM.2.5 Concentration Value (PPB)") +
  xlab("Values") +
  ylab("Frequency")
```


```{r Log transform PM2.5 values the graph again}
main_df$log_pm2.5 <- log(main_df$PM.2.5)
ggplot(main_df, aes(x = log_pm2.5)) +
  geom_histogram(bins = 10, fill = "blue", color = "black") +
  ggtitle("Distribution of Log Transformed PM.2.5 Concentration Value (PPB)") +
  xlab("Values") +
  ylab("Frequency")
```
> Since the original data of PM2.5 displays some level of left-skewness, I attempted to log-transform the PM2.5 values, but the distribution stil displays a left-skewness. We can experiment this with regressions later too.



```{r Distrbution of NO2 Value }
main_df$NO2 <- as.numeric(main_df$NO2)

ggplot(main_df, aes(x = NO2)) +
  geom_histogram(bins = 10, fill = "purple", color = "black") +
  ggtitle("Distribution of NO2 Concentration Value (PPB)") +
  xlab("Values") +
  ylab("Frequency")
```



```{r}
main_df$Ozone_ppm <- as.numeric(main_df$Ozone_ppm)
ggplot(main_df, aes(x = Ozone_ppm)) +
  geom_histogram(bins = 10, fill = "orange", color = "black") +
  ggtitle("Distribution of Ozone Concentration Value (PPM)") +
  xlab("Values") +
  ylab("Frequency")
```
> The concentration values (parts per billion) of NO2 and Ozone (parts per million) are closer to normal distribution. 
> The two values in Ozone concentration histograms may appear to be outliers according to the histogram grouping, but we decided not to remove them because of small range of values.


## Independent Variables


