---
title: "Summary_Statistics"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(maps)
library(skimr)
library(psych)
library(moments)
library(nortest)
```

```{r load-data}
# Load data
data <- read_csv("Data and Raw Data Wrangling/Combined Final Dataframe.csv")
```

# Summary Statistics

```{r summary-stats}
# Summary statistics
summary(data)
skim(data)
```

# Exploratory Data Visuals

```{r State Map}
data_clean <- data %>%
  mutate(region = tolower(State))

states_map <- map_data("state")

# Merge map data
map_data_joined <- left_join(states_map, data_clean, by = "region")

# Plot
ggplot(map_data_joined, aes(x = long, y = lat, group = group, fill = EV_ratio)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  labs(title = "EVs per Capita by State", fill = "EVs per Capita") +
  theme_void()
```



```{r Maps of Urbanization and EV Ratio by States}
library(ggplot2)
library(dplyr)
library(maps)
library(gridExtra)
library(moments)    # For skewness & kurtosis
library(nortest)    # For normality tests

# Prepare data
data_clean <- data %>% mutate(region = tolower(State))
states_map <- map_data("state")
map_data_joined <- left_join(states_map, data_clean, by = "region")

# Urbanization map
urban_map <- ggplot(map_data_joined, aes(x = long, y = lat, group = group, fill = Urbanization)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  scale_fill_continuous(name = "Urban %", low = "#deebf7", high = "#08519c") +
  labs(title = "Urbanization by State") +
  theme_void()

# EVs per capita map
ev_map <- ggplot(map_data_joined, aes(x = long, y = lat, group = group, fill = EV_ratio)) +
  geom_polygon(color = "white") +
  coord_fixed(1.3) +
  scale_fill_continuous(name = "EVs per Capita", low = "#fee5d9", high = "#a50f15") +
  labs(title = "EVs per Capita by State") +
  theme_void()

# Display side-by-side
gridExtra::grid.arrange(urban_map, ev_map, ncol = 2)
```

```{r Scatterplot Urbanization and EV ratio}
library(ggrepel)

ggplot(data, aes(x = Urbanization, y = EV_ratio, label = State)) +
  geom_point(color = "#1f77b4", size = 3) +
  geom_text_repel(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Urbanization vs EVs per Capita by State",
       x = "% Urban Population",
       y = "EVs per Capita") +
  theme_minimal()
```


## Dependent Variables Histograms 

```{r PM2.5 Value Distribution}
ggplot(data, aes(x = PM.2.5)) +
  geom_histogram(bins = 10, fill = "blue", color = "black") +
  ggtitle("Distribution of PM.2.5 Concentration Value (PPB)") +
  xlab("Values") +
  ylab("Frequency")
```


```{r Log transform PM2.5 values the graph again}
data$log_pm2.5 <- log(data$PM.2.5)
ggplot(data, aes(x = log_pm2.5)) +
  geom_histogram(bins = 10, fill = "blue", color = "black") +
  ggtitle("Distribution of Log Transformed PM.2.5 Concentration Value (PPB)") +
  xlab("Values") +
  ylab("Frequency")
```
> Since the original data of PM2.5 displays some level of left-skewness, I attempted to log-transform the PM2.5 values, but the distribution stil displays a left-skewness. We can experiment this with regressions later too.



```{r Distrbution of NO2 Concentrations}
data$NO2 <- as.numeric(data$NO2)

ggplot(data, aes(x = NO2)) +
  geom_histogram(bins = 10, fill = "purple", color = "black") +
  ggtitle("Distribution of NO2 Concentration Value (PPB)") +
  xlab("Values") +
  ylab("Frequency")
```



```{r}
data$Ozone_ppm <- as.numeric(data$Ozone_ppm)
ggplot(data, aes(x = Ozone_ppm)) +
  geom_histogram(bins = 10, fill = "orange", color = "black") +
  ggtitle("Distribution of Ozone Concentration Value (PPM)") +
  xlab("Values") +
  ylab("Frequency")
```
> The concentration values (parts per billion) of NO2 and Ozone (parts per million) are closer to normal distribution. 
> The two values in Ozone concentration histograms may appear to be outliers according to the histogram grouping, but we decided not to remove them because of small range of values.

## Independent Variables Histograms
```{r EV Registeration Ratio}
data$EV_ratio <- as.numeric(data$EV_ratio)
ggplot(data, aes(x = EV_ratio)) +
  geom_histogram(bins = 10, fill = "blue", color = "black") +
  ggtitle("Distribution of EV Per Capita") +
  xlab("Values") +
  ylab("Frequency")
```
> EV per capita ratio is also right skewed.


```{r Temperature}
data$Temp_F <- as.numeric(data$Temp_F)
ggplot(data, aes(x = Temp_F)) +
  geom_histogram(bins = 10, fill = "purple", color = "black") +
  ggtitle("Distribution of 2022 Average Temperatures Across States") +
  xlab("Values") +
  ylab("Frequency")
```


```{r Precipitation}
data$Precip_in <- as.numeric(data$Precip_in)
ggplot(data, aes(x = Precip_in)) +
  geom_histogram(bins = 10, fill = "orange", color = "black") +
  ggtitle("Distribution of 2022 Average Precipitation (inches) Across States") +
  xlab("Values") +
  ylab("Frequency")
```

## Correlation Coefficients between Variables
```{r}

```


# Regressions

## Regression for PM2.5 (PPB)

```{r Original PM2.5}
model_pm25 <- lm(PM.2.5 ~ EV_ratio + Urbanization + Temp_F + Precip_in, data = data)
summary(model_pm25)
```
```{r Log-tranform non-significant independent variables}
data$log_Urbanization <- log(data$Urbanization)
data$log_EV_ratio <- log(data$EV_ratio)
data$log_Precip_in <- log(data$Precip_in)
```


```{r Log-transformed regression of PM2.5}
# re-run model after log-trans
model_logpm25 <- lm(log_pm2.5 ~ log_EV_ratio + log_Urbanization + Temp_F + log_Precip_in, data = data)
summary(model_logpm25)
```
> The log-transformed model made the p-values lower for each variable and the overall model, but it did change the non-significant variables to significant.


## Regression for NO2 (PPB)




## Regression for Ozone -- O3 (PPM)
### Data exploration 
```{r}
skim(data$Ozone_ppm)
skim(data$Population_2022)
colnames(data)

ggplot(data, aes(x = Ozone_ppm)) +
  geom_histogram() +
  labs(title = "Ozone Plot",
       x = "Value",
       y = "Frequency") +
  theme_minimal()

ggplot(data, aes(x = Urbanization)) +
  geom_histogram() +
  labs(title = "Urbanization",
       x = "Value",
       y = "Frequency") +
  theme_minimal()

ggplot(data, aes(x = Temp_F)) +
  geom_histogram() +
  labs(title = "Temperature",
       x = "Value",
       y = "Frequency") +
  theme_minimal()

ggplot(data, aes(x = EV_ratio)) +
  geom_histogram() +
  labs(title = "EV ratio",
       x = "Value",
       y = "Frequency") +
  theme_minimal()
```
> All variables are skewed. Trying to log transform them 

### Testing normality
```{r}
# Normality test function 

test_normality <- function(data, var_name) {
  clean_var <- na.omit(data[[var_name]])
  cat("\n=== Normality Test for", var_name, "===\n")
  # Descriptive statistics
  cat("\nDescriptive Statistics:\n")
  print(describe(clean_var))
  
  # Skewness and kurtosis
  cat("\nSkewness:", skewness(clean_var))
  cat("\nKurtosis:", kurtosis(clean_var), "\n")
  
  # Shapiro-Wilk test (for samples ≤ 5000)
  if(length(clean_var) <= 5000) {
    shapiro_test <- shapiro.test(clean_var)
    cat("\nShapiro-Wilk Test:\n")
    print(shapiro_test)
  } else {
    cat("\nSample too large for Shapiro-Wilk test\n")
  }

  # Anderson-Darling test
  ad_test <- ad.test(clean_var)
  cat("\nAnderson-Darling Test:\n")
  print(ad_test)
}

# Normality test 
test_normality(data, "Ozone_ppm") # highly skewed, need log transform 
test_normality(data, "Urbanization") # no need for log transform 
test_normality(data, "Temp_F") # normally distributed, no log needed
test_normality(data, "Precip_in") # skewed, need log transform
test_normality(data, "EV_ratio") # skewed, need log transformation

# Log transformation 
log_ozone <- log(data$Ozone_ppm)
log_precip <- log(data$Precip_in)
log_EV_ratio <- log(data$EV_ratio)

# Regression 
ozone_model <- lm(log_ozone ~ log_precip + log_EV_ratio + Temp_F + Urbanization, data = data)
summary(ozone_model)

#check residuals
plot(ozone_model)
```


