---
title: "Time series analysis"
author: "Faustin Kambale"
date: "2025-03-19"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Insert required packages

```{r}
library (here)
library(dplyr)
library(readxl)
library (ggplot2)
library(readr)
library(tidyr)
library(dplyr)
library(lubridate)
library(forecast)   
library(tsibble)    
library(feasts) 
library(fabletools)
library(tseries)
```

# Getting the dataset

```{r}
getwd()
dataset <- read.csv("cleaned_dataset.csv", stringsAsFactors = FALSE)
str(dataset)
dataset$year <- as.Date(paste0(dataset$year, "-01-01"), format = "%Y-%m-%d")
dataset$year <- year(dataset$year) # Extracting only the year
head(dataset,5)
```

# Step 1: Create Time Series Objects for All 10 States

```{r}
# Extract gross total emissions for each state
alabama <- dataset$alab_gross.total
alaska <- dataset$alaska_gross.total
arizona <- dataset$arizona_gross.total
arkansas <- dataset$arkansas_gross.total
california <- dataset$california_gross.total
virginia <- dataset$virginia_gross.total
washington <- dataset$washington_gross.total
west_virginia <- dataset$west_virginia_gross.total
wisconsin <- dataset$wisconsin_gross.total
wyoming <- dataset$wyoming_gross.total

# Create time series objects for each state (annual data)
ts_alabama <- ts(alabama, start = 1991, frequency = 1)
ts_alaska <- ts(alaska, start = 1991, frequency = 1)
ts_arizona <- ts(arizona, start = 1991, frequency = 1)
ts_arkansas <- ts(arkansas, start = 1991, frequency = 1)
ts_california <- ts(california, start = 1991, frequency = 1)
ts_virginia <- ts(virginia, start = 1991, frequency = 1)
ts_washington <- ts(washington, start = 1991, frequency = 1)
ts_west_virginia <- ts(west_virginia, start = 1991, frequency = 1)
ts_wisconsin <- ts(wisconsin, start = 1991, frequency = 1)
ts_wyoming <- ts(wyoming, start = 1991, frequency = 1)

# Combine all time series into a single matrix for easier comparison (optional)
ts_all_states <- cbind(ts_alabama, ts_alaska, ts_arizona, ts_arkansas, 
                       ts_california, ts_virginia, ts_washington, 
                       ts_west_virginia, ts_wisconsin, ts_wyoming)
```

# Step 2: Explore the timeseries 

```{r}
# Rename columns for clarity
colnames(ts_all_states) <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", 
                            "Virginia", "Washington", "West Virginia", "Wisconsin", "Wyoming")
# Plotting the state's emissions
matplot(ts_all_states, type = "l", lty = 1, col = 1:10, 
        main = "Cross sector Gross Total Emmission by State (1991-2023)", 
        ylab = "Gross Total (Units)", xlab = "Year")
legend("topleft", legend = colnames(ts_all_states), col = 1:10, lty = 1, cex = 0.8)

## or the below plot

plot(ts_all_states) # Plotting all TS 
hist (ts_alabama) #Can plot state by state in a small multiple (to do later)

```
From this plot, we can notice that, among all the states, California pollutes the most.

## Enchanced plot (should check other plot format from the TSA class. This one is not easy to compare.)

```{r}
# Convert the combined time series to a data frame for ggplot2
ts_df <- as.data.frame(ts_all_states)
ts_df$Year <- 1991:2023

# Reshape to long format
ts_long <- ts_df %>%
  pivot_longer(cols = -Year, names_to = "State", values_to = "Gross_Total")

# Plot with ggplot2 (faceted)
ggplot(ts_long, aes(x = Year, y = Gross_Total)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") +  # Separate panels with individual y-scales
  labs(title = "Cross sector Gross Total Emmission by State (1991-2023)", 
       x = "Year", y = "Gross Total (Units)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Data visualization 

```{r}
matplot(ts_all_states, type = "l", lty = 1, col = 1:10, 
        main = "Gross Total Emmission by State (1991-2023)", 
        ylab = "Gross Total (Units)", xlab = "Year")
legend("topleft", legend = colnames(ts_all_states), col = 1:10, lty = 1, cex = 0.8)
```

```{r}
ggplot(ts_long, aes(x = Year, y = Gross_Total)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") + # or try scales = "free_x" to see the #ce
  labs(title = "Cross state Gross Total Emission (1991-2023)", 
       x = "Year", y = "Gross Total (Units)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Observation: Look for trends (upward, downward, or stable) in each state. Note that Alabama and Alaska appear identical, indicating a potential data error. California has the highest trend.

# Step 3: Decomposition

### Since the data is annual (frequency = 1), I use a simple moving average to estimate trends for each state.

```{r}
trend_all <- apply(ts_all_states, 2, function(x) {
  ma(x, order = 3, centre = TRUE)  # 3-year moving average
})

# Plot trends for all states
matplot(trend_all, type = "l", lty = 1, col = 1:10, 
        main = "3-Year Moving Average Trends by State (1991-2023)", 
        ylab = "Smoothed Gross Total", xlab = "Year")
legend("topleft", legend = colnames(ts_all_states), col = 1:10, lty = 1, cex = 0.8)
```

### Stationarity Test for each state using the ADF test
```{r}
# Perform ADF test for each state
adf_results <- apply(ts_all_states, 2, function(x) {
  adf.test(x, alternative = "stationary")$p.value
})

# Display p-values
adf_df <- data.frame(State = colnames(ts_all_states), ADF_p_value = adf_results)
print(adf_df)
```

Interpretation: If p-value > 0.05, the series is non-stationary (common for energy consumption data with trends). We’ll difference non-stationary series next.

### Differencing for any stationary series (first difference)

```{r}
ts_all_diff <- apply(ts_all_states, 2, diff)
```

### Re-run ADF test on differenced data
```{r}
adf_diff_results <- apply(ts_all_diff, 2, function(x) {
  adf.test(x, alternative = "stationary")$p.value
})

# Display results (corrected: use all state names)
adf_diff_df <- data.frame(State = colnames(ts_all_states), ADF_p_value_Diff = adf_diff_results)
print(adf_diff_df)

# Plot differenced series
matplot(ts_all_diff, type = "l", lty = 1, col = 1:10, 
        main = "Differenced Gross Total Emmissions by State", 
        ylab = "Difference", xlab = "Year")
legend("bottomleft", legend = colnames(ts_all_states), col = 1:10, lty = 1, cex = 0.8)
```

Intererpretation: If p-value < 0.05, the differenced series is stationary.

# Step 4:  Fit a Time Series Model

```{r}
# Fit ARIMA models to all states
arima_models <- apply(ts_all_states, 2, function(x) {
  auto.arima(x, stepwise = TRUE, approximation = TRUE)
})
```

## Summarize each model (check notes for a better way to present them)
```{r}

for (state in names(arima_models)) {
  cat("\nState:", state, "\n")
  print(summary(arima_models[[state]]))
}

# Extract fitted values
fitted_values <- sapply(arima_models, fitted)

# Plot observed vs fitted
matplot(ts_all_states, type = "l", lty = 1, col = 1:10, 
        main = "Observed vs Fitted Gross Total Energy Consumption", 
        ylab = "Gross Total", xlab = "Year")
matlines(fitted_values, lty = 2, col = 1:10)
legend("topleft", legend = colnames(ts_all_states), col = 1:10, lty = 1, cex = 0.8)
legend("topright", legend = "Fitted", lty = 2, col = "black")

# Check residuals
par(mfrow = c(5, 2))
for (state in names(arima_models)) {
  checkresiduals(arima_models[[state]], main = paste("Residuals for", state))
}
par(mfrow = c(1, 1))
```
Interpretation???
Good Fit: Residuals should be white noise (Ljung-Box p-value > 0.05)

# Step 5: Forecast Future Values (5 years (2024–2028) for all states.)

```{r}
forecasts <- lapply(arima_models, forecast, h = 5)

# Extract point forecasts and confidence intervals
forecast_points <- t(sapply(forecasts, function(x) x$mean))
forecast_lower <- t(sapply(forecasts, function(x) x$lower[, "95%"]))
forecast_upper <- t(sapply(forecasts, function(x) x$upper[, "95%"]))

# Plot forecasts (one plot per state)
par(mfrow = c(5, 2))
for (state in names(forecasts)) {
  plot(forecasts[[state]], main = paste("Forecast for", state, "(2024-2028)"), 
       ylab = "Gross Total", xlab = "Year")
}
par(mfrow = c(1, 1))

# Alternative: ggplot2 faceted plot
forecast_df <- data.frame(
  Year = rep(2024:2028, times = 10),
  State = rep(colnames(ts_all_states), each = 5),
  Forecast = as.vector(t(forecast_points)),
  Lower_95 = as.vector(t(forecast_lower)),
  Upper_95 = as.vector(t(forecast_upper))
)

ggplot(forecast_df, aes(x = Year, y = Forecast)) +
  geom_line() +
  geom_ribbon(aes(ymin = Lower_95, ymax = Upper_95), alpha = 0.2) +
  facet_wrap(~ State, scales = "free_y") +
  labs(title = "5-Year Forecast of Gross Total Energy Consumption by State", 
       x = "Year", y = "Gross Total (Units)") +
  theme_minimal()
```

# Next Step 6: Interpret and Report Results
Hints : Trends: Summarize trends from Step 4 (e.g., California increases, Wyoming fluctuates).
 Stationarity: Note which states required differencing (from ADF results).
 Models: List ARIMA parameters (e.g., Alabama: ARIMA(1,1,0), California: ARIMA(2,1,1)).
 Forecasts: Highlight key predictions (e.g., "California’s gross total is projected to reach X by 2028 with a 95% CI of [Y, Z]").
 Example report snippet:
"Alabama and Alaska data appear duplicated, requiring correction. Arizona shows a steady increase, while Wyoming remains stable. ARIMA models fit well for most states, with forecasts indicating continued growth in California and Virginia."

# Step 7 : Save the Work
```{r}
# Save the script
# File > Save As > "time_series_analysis_all_states.R"

# Export forecast plot
ggsave("all_states_forecast.png", width = 12, height = 8)

# Export forecast values to CSV
write.csv(forecast_df, "all_states_forecast_values.csv", row.names = FALSE)
```

TO BE CONTINUED  